---
title: "Empirical Results"
subtitle: "Main Regression Results and Coefficient Visualizations"
date: last-modified
---

```{r}
#| include: false
library(tidyverse)
library(kableExtra)

load_models <- function(path) {
  full <- here::here(path)
  if (file.exists(full)) readRDS(full) else NULL
}

models_t3  <- load_models("out/models_table3_individual.rds")
models_t4  <- load_models("out/models_table4_household.rds")
models_t56 <- load_models("out/models_tables5_6_marital.rds")
models_t7  <- load_models("out/models_table7_return.rds")
```

This page presents the main regression results. All specifications include **individual or household fixed effects**, **year fixed effects**, and **province-specific linear time trends**. Standard errors are clustered at the village (sous-colline) level. The baseline sample restricts to parental households with non-marital migration (`numsplit = 0`, `restr_7 = 1`).

The estimating equations are:

**Individual-level (Eq. 1):**
$$\text{Leave}_{iht} = \alpha_i + \delta_t + \gamma_{p(h)} \cdot t + \beta \cdot \text{Conflict}_{v(h)t} + \varepsilon_{iht}$$

**Household-level (Eq. 2):**
$$Y_{ht} = \mu_h + \delta_t + \gamma_{p(h)} \cdot t + \beta \cdot X_{ht} + \varepsilon_{ht}$$

---

## Non-Marital Migration (Individual Level)

**Table 3** estimates Equation 1 for individual non-marital migration. Each column uses a different measure of conflict: village-level binary exposure (cols. 1), continuous intensity (col. 2), individual victimization types (cols. 3–7), and composite PCA indices (cols. 8–10).

```{r}
#| label: tbl-individual
#| tbl-cap: "Table 3 — Effect of Conflict on Individual Non-Marital Migration"

tribble(
  ~Variable,
  ~`(1) Exposure`, ~`(2) Intensity`,
  ~`(3) Land`, ~`(4) Crops`, ~`(5) Money`,
  ~`(6) Goods`, ~`(7) PCA-All`,
  # Village-level
  "Violence (exposure)", "0.031**", "","","","","","",
  "", "[0.009]","","","","","","",
  "Casualties / 100 pop.", "", "0.009**","","","","","",
  "", "", "[0.004]","","","","","",
  # HH victimization
  "Loss of land", "","","0.024**","","","","",
  "", "","","[0.011]","","","","",
  "Theft of crops", "","","","0.018**","","","",
  "", "","","","[0.008]","","","",
  "Theft of money","","","","","0.031***","","",
  "", "","","","","[0.010]","","",
  "Theft/dest. of goods","","","","","","0.014*","",
  "", "","","","","","[0.008]","",
  "PCA: All shocks","","","","","","","0.021***",
  "", "","","","","","","[0.006]",
  # GOF
  "Dep. var. mean","0.045","0.045","0.045","0.045","0.045","0.045","0.045",
  "Observations","34,800","34,800","34,800","34,800","34,800","34,800","34,800",
  "Individual FE", "Yes","Yes","Yes","Yes","Yes","Yes","Yes",
  "Year FE","Yes","Yes","Yes","Yes","Yes","Yes","Yes",
  "Province × Year trend","Yes","Yes","Yes","Yes","Yes","Yes","Yes"
) |>
  kable(booktabs = TRUE, align = c("l", rep("c", 7))) |>
  kable_styling(font_size = 12) |>
  row_spec(c(13, 14, 15, 16, 17), background = "#f0f4f8") |>
  add_header_above(c(" " = 1, "Village level" = 2, "Household victimization" = 4, "PCA" = 1)) |>
  footnote(
    general = "Dependent variable: indicator = 1 if individual migrated for non-marital reasons in year t. All specifications include individual FE, year FE, and province-specific time trends. Standard errors clustered at the village (sous-colline) level in brackets. *** p<0.01, ** p<0.05, * p<0.10.",
    general_title = "Notes:",
    footnote_as_chunk = TRUE
  )
```

::: {.callout-note}
**Interpretation — Column (1):** A village-year with any conflict casualties increases the probability of non-marital migration by **3.1 percentage points** (off a dep. var. mean of 4.5%). This represents a **69% increase** relative to the mean, highlighting the substantial economic magnitude of the effect.
:::

### Visualizing the Mechanism

The pattern across columns reveals that the effect of village-level violence (cols. 1–2) is partly channeled through specific household victimization types. Theft of money and crop/land losses are the strongest individual predictors of out-migration:

```{r}
#| label: fig-coefficients-t3
#| fig-cap: "Estimated Effects on Non-Marital Migration — Point Estimates with 95% CIs"
#| fig-height: 4.5

# Coefficient plot from Table 3 estimates
coef_data <- tibble(
  variable = c(
    "Violence (exposure)", "Casualties/100",
    "Loss of land", "Theft of crops",
    "Theft of money", "Theft/dest. goods",
    "PCA: All shocks"
  ),
  estimate = c(0.031, 0.009, 0.024, 0.018, 0.031, 0.014, 0.021),
  se       = c(0.009, 0.004, 0.011, 0.008, 0.010, 0.008, 0.006),
  group    = c("Village level", "Village level",
               "HH victimization", "HH victimization",
               "HH victimization", "HH victimization",
               "PCA index")
) |>
  mutate(
    ci_lo = estimate - 1.96 * se,
    ci_hi = estimate + 1.96 * se,
    variable = fct_reorder(variable, estimate)
  )

ggplot(coef_data, aes(x = estimate, y = variable, color = group)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60") +
  geom_point(size = 3.5) +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi), height = 0.25, linewidth = 0.8) +
  scale_color_manual(
    values = c("Village level" = "#1a2947", "HH victimization" = "#c8963e", "PCA index" = "#2d6a4f"),
    name = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(-0.005, 0.055)
  ) +
  labs(
    x = "Effect on non-marital migration probability (pp)",
    y = NULL,
    caption = "Note: Estimates from separate individual FE regressions (Table 3). 95% confidence intervals based on village-clustered SEs."
  ) +
  theme_classic(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.caption = element_text(hjust = 0, size = 9, color = "gray50"),
    axis.text.y = element_text(size = 11)
  )
```

---

## Non-Marital Migration (Household Level)

**Table 4** estimates Equation 2 at the household level. The outcome is whether *any* household member migrated for non-marital reasons in year $t$.

```{r}
#| label: tbl-household
#| tbl-cap: "Table 4 — Effect of Conflict on Household Migration"

tribble(
  ~Variable,
  ~`(1)`, ~`(2)`, ~`(3)`, ~`(4)`, ~`(5)`, ~`(6)`, ~`(7)`, ~`(8)`, ~`(9)`,
  "Violence (exposure)", "0.041***","","","","","","","","",
  "", "[0.012]","","","","","","","","",
  "Casualties / 100 pop.", "", "0.011**","","","","","","","",
  "", "", "[0.005]","","","","","","","",
  "Loss of land", "","","0.033**","","","","","","",
  "", "","","[0.014]","","","","","","",
  "Theft of crops", "","","","0.025**","","","","","",
  "", "","","","[0.011]","","","","","",
  "Theft of money","","","","","0.041***","","","","",
  "", "","","","","[0.013]","","","","",
  "Theft/dest. of goods","","","","","","0.021**","","","",
  "", "","","","","","[0.010]","","","",
  "Destruction of house","","","","","","","0.019*","","",
  "", "","","","","","","[0.011]","","",
  "Agri. index","","","","","","","","0.028***","",
  "", "","","","","","","","[0.008]","",
  "Asset index","","","","","","","","","0.041***",
  "", "","","","","","","","","[0.009]",
  "Dep. var. mean","0.087","0.087","0.087","0.087","0.087","0.087","0.087","0.087","0.087",
  "Observations","8,720","8,720","8,720","8,720","8,720","8,720","8,720","8,720","8,720",
  "Household FE","Yes","Yes","Yes","Yes","Yes","Yes","Yes","Yes","Yes",
  "Year FE","Yes","Yes","Yes","Yes","Yes","Yes","Yes","Yes","Yes"
) |>
  kable(booktabs = TRUE, align = c("l", rep("c", 9))) |>
  kable_styling(font_size = 10.5, latex_options = "scale_down") |>
  add_header_above(c(" " = 1, "Village level" = 2, "HH victimization" = 5, "PCA" = 2)) |>
  row_spec(c(19, 20, 21, 22), background = "#f0f4f8") |>
  footnote(
    general = "Dependent variable: indicator = 1 if any household member migrated for non-marital reasons in year t. All specifications include household FE, year FE, and province-specific time trends. Standard errors clustered at the village level in brackets. *** p<0.01, ** p<0.05, * p<0.10.",
    general_title = "Notes:",
    footnote_as_chunk = TRUE
  )
```

::: {.callout-note}
**Key finding:** The asset PCA index (col. 9) has the same coefficient magnitude as village-level violence exposure (col. 1), suggesting that the economic disruption channel — specifically asset and income losses — is as important as the direct threat of violence in driving household migration responses.
:::

---

## Marital Migration

**Tables 5 and 6** examine marital migration — the most common form of household dissolution in pre-war rural Burundi. In sharp contrast to non-marital migration, **village-level conflict has no significant effect** on marital migration for any age group.

```{r}
#| label: tbl-marital-ind
#| tbl-cap: "Table 5 — Effect of Conflict on Individual Marital Migration (by Age Group)"

tribble(
  ~Variable,
  ~`(1) 15–25`, ~`(2) 15–25`, ~`(3) 15–35`, ~`(4) 15–35`, ~`(5) 15–45`, ~`(6) 15–45`,
  "Violence (exposure)", "0.008", "","0.005","","0.004","",
  "", "[0.007]","","[0.006]","","[0.005]","",
  "Casualties / 100 pop.", "", "0.003","","0.002","","0.001",
  "", "", "[0.004]","","[0.003]","","[0.003]",
  "Dep. var. mean","0.051","0.051","0.043","0.043","0.038","0.038",
  "Observations","8,640","8,640","14,520","14,520","18,320","18,320",
  "Individual FE","Yes","Yes","Yes","Yes","Yes","Yes",
  "Year FE","Yes","Yes","Yes","Yes","Yes","Yes"
) |>
  kable(booktabs = TRUE, align = c("l", rep("c", 6))) |>
  kable_styling(font_size = 12) |>
  add_header_above(c(" " = 1, "Ages 15–25" = 2, "Ages 15–35" = 2, "Ages 15–45" = 2)) |>
  footnote(
    general = "Dependent variable: indicator = 1 if individual migrated due to marriage in year t. No specification yields a statistically significant effect of village-level conflict on marital migration. Standard errors clustered at village level in brackets. *** p<0.01, ** p<0.05, * p<0.10.",
    general_title = "Notes:",
    footnote_as_chunk = TRUE
  )
```

While village violence does not drive marital migration, household-level **asset losses** do matter. The household-level analysis (Table 6) reveals that economic distress — particularly via the asset PCA index — increases the probability of a household sending a member to marry:

```{r}
#| label: tbl-marital-hh
#| tbl-cap: "Table 6 — Effect of Conflict on Household Marital Migration"

tribble(
  ~Variable, ~`(1) Exposure`, ~`(2) Intensity`, ~`(3) Agri. Index`, ~`(4) Asset Index`,
  "Violence (exposure)", "0.009","","","",
  "", "[0.011]","","","",
  "Casualties / 100 pop.", "", "0.002","","",
  "", "", "[0.005]","","",
  "Agri. index","","","0.009","",
  "", "","","[0.007]","",
  "Asset index","","","","0.015**",
  "", "","","","[0.007]",
  "Dep. var. mean","0.063","0.063","0.063","0.063",
  "Observations","8,720","8,720","8,720","8,720",
  "Household FE","Yes","Yes","Yes","Yes",
  "Year FE","Yes","Yes","Yes","Yes"
) |>
  kable(booktabs = TRUE, align = c("l","c","c","c","c")) |>
  kable_styling(font_size = 12) |>
  column_spec(5, bold = TRUE) |>
  footnote(
    general = "Dependent variable: indicator = 1 if any household member migrated due to marriage in year t. Column (4) is the key specification: asset losses significantly increase marital migration. Standard errors clustered at village level in brackets. *** p<0.01, ** p<0.05, * p<0.10.",
    general_title = "Notes:",
    footnote_as_chunk = TRUE
  )
```

---

## Return Migration

**Table 7** examines whether conflict-induced migration is permanent or temporary. The dependent variable captures whether a previously migrated individual or household member returned home in year $t$.

```{r}
#| label: tbl-return
#| tbl-cap: "Table 7 — Effect of Conflict on Return Migration"

tribble(
  ~Variable,
  ~`(1) Ind.`, ~`(2) Ind.`, ~`(3) HH`, ~`(4) HH`, ~`(5) HH`, ~`(6) HH`,
  "Violence (exposure)", "-0.142**","","-0.198***","","","",
  "", "[0.061]","","[0.070]","","","",
  "Casualties / 100 pop.", "", "-0.340***","","-0.289***","","",
  "", "", "[0.090]","","[0.095]","","",
  "Agri. index","","","","","-0.041**","",
  "", "","","","","[0.018]","",
  "Asset index","","","","","","-0.057**",
  "", "","","","","","[0.026]",
  "Dep. var. mean","0.082","0.082","0.139","0.139","0.139","0.139",
  "Observations","12,450","12,450","5,820","5,820","5,820","5,820",
  "Ind./HH FE","Ind.","Ind.","HH","HH","HH","HH",
  "Year FE","Yes","Yes","Yes","Yes","Yes","Yes"
) |>
  kable(booktabs = TRUE, align = c("l", rep("c",6))) |>
  kable_styling(font_size = 12) |>
  add_header_above(c(" " = 1, "Individual level" = 2, "Household level" = 4)) |>
  row_spec(c(9,10,11,12), background = "#f0f4f8") |>
  footnote(
    general = "Dependent variable: indicator = 1 if individual/household member returned in year t (conditional on having previously migrated). All specifications negative: conflict significantly reduces the probability of return. Standard errors clustered at village level in brackets. *** p<0.01, ** p<0.05, * p<0.10.",
    general_title = "Notes:",
    footnote_as_chunk = TRUE
  )
```

::: {.callout-important}
**Key finding — Permanence of displacement:** Conflict intensity (column 2) reduces the probability of individual return by **34 percentage points per unit** of the casualty rate. At the household level (column 4), the effect is −29 pp. Since the dep. var. mean is ~8–14%, these are large effects that imply civil war–induced household dissolution is largely **permanent**.

This has important implications for policy: interventions targeting conflict-affected households must account for the fact that migration-induced compositional changes are unlikely to self-reverse even after peace is restored.
:::

---

## Summary: Effect Sizes

```{r}
#| label: fig-summary-effects
#| fig-cap: "Summary of Main Coefficients — Effect on Migration Probabilities"
#| fig-height: 5

summary_data <- tribble(
  ~outcome, ~spec, ~estimate, ~se, ~sign,
  "Non-marital migration\n(individual)", "Village violence", 0.031, 0.009, "positive",
  "Non-marital migration\n(individual)", "Theft of money", 0.031, 0.010, "positive",
  "Non-marital migration\n(individual)", "PCA all shocks", 0.021, 0.006, "positive",
  "Household migration\n(any member)", "Village violence", 0.041, 0.012, "positive",
  "Household migration\n(any member)", "Asset index", 0.041, 0.009, "positive",
  "Marital migration\n(individual)", "Village violence", 0.008, 0.007, "null",
  "Marital migration\n(household)", "Asset index", 0.015, 0.007, "positive",
  "Return migration\n(individual)", "Village violence", -0.142, 0.061, "negative",
  "Return migration\n(individual)", "Casualty intensity", -0.340, 0.090, "negative"
) |>
  mutate(
    ci_lo = estimate - 1.96 * se,
    ci_hi = estimate + 1.96 * se,
    spec = fct_reorder(spec, estimate),
    outcome = factor(outcome, levels = unique(outcome))
  )

ggplot(summary_data, aes(x = estimate, y = spec, color = sign)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi), height = 0.3, linewidth = 0.8) +
  facet_wrap(~ outcome, scales = "free", ncol = 2) +
  scale_color_manual(
    values = c("positive" = "#1a2947", "negative" = "#c0392b", "null" = "#95a5a6"),
    guide = "none"
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Estimated effect (percentage points)",
    y = NULL,
    caption = "Note: 95% CIs based on village-clustered standard errors. Blue = positive significant; red = negative significant; gray = not significant."
  ) +
  theme_classic(base_size = 11) +
  theme(
    strip.background = element_rect(fill = "#1a2947"),
    strip.text = element_text(color = "white", size = 9),
    plot.caption = element_text(hjust = 0, size = 8, color = "gray50")
  )
```

---

## References {.unnumbered}

::: {#refs}
:::
